\documentclass{scrartcl}
\usepackage[utf8x]{inputenc}
\usepackage[USenglish]{babel}
\usepackage{url}
\usepackage{todonotes}

\title{A Language for the Specification and Efficient Implementation
  of Type Systems}
\subtitle{Master's Thesis}
\begin{document}
\maketitle
\todo[inline]{What about the addition ``static'' in the title}
\todo[inline]{Write some essay-like motivation}
\todo[inline]{More about extending a given type system}
\todo[inline]{How do theorem proves play a role here}

Type systems are means to put focus on meaningful programs. In other
words they help the programmer to catch errors by helping building
abstractions and by enforcing those abstractions thorough the
program. We are dealing with static type systems, i.e. abstractions
are enforced before the execution. Those type systems help to reduce
development time and to reveal errors earlier. \todo{Cite Stefan
  Hanenberg}

Static type systems are typically integrated into the compiler of the
programming language. \todo{Proof this} This makes it hard to extend a
given type system or to implement a type system for an (embedded)
domain specific language. In addition the implementation of a type
system is time expensive and error prone.

The goal of this thesis is to design a language for the specification
of static type systems in the style of natural deduction and to
generate based on such specifications an efficient implementation.

\section{Project Objectives}

\subsection{Core:}
\begin{description}
\item[{\bf A: Domain Specific Language to describe Type Systems}]
\item[{\bf B: Generation of Type Checkers}]
\item[{\bf C: Interface to Automated Theorem Provers}]
\end{description}

\subsection{Extensions:}

Additionally to the core objectives, the following objectives could be
pursued.

\begin{itemize}
\item Trivial attempts to soundness checks?
\item Does it scale to for example security type systems?
\end{itemize}

\section{Main Activities}
\begin{description}
\item[{\bf A}] includes
  \begin{itemize}
    \item
  \end{itemize}

\item[{\bf B}] includes
  \begin{itemize}
    \item
  \end{itemize}

\item[{\bf C}] includes
  \begin{itemize}
  \item 
  \end{itemize}

\item[{\bf D}] includes
  \begin{itemize}
  \item
  \end{itemize}
\end{description}

\section{Deliverables}
The master's thesis shall include

\begin{itemize}
\item
\end{itemize}

After completing the thesis, a talk shall present the main results of the thesis.

\section{Prerequisites}

\begin{itemize}
\item 
\end{itemize}

\section{Supervision}
Dr. Sebastian Erdweg \\
(Software Technology Group)

\bibliographystyle{alpha}
\small
\bibliography{literature}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
